---
title: "SISMID Tidyverse Module 3 Code-along"
output: html_document
date: "2024-07-18"
---

```{r load packages}
library(tidyverse)
library(nycflights13)

flights <- nycflights13::flights
weather <- nycflights13::weather
```

## FILTER

Explain what each of line of code is accomplishing
```{r filter 1}
jfk_flights <- filter(flights, origin == "JFK")
nyc_am_flights <- filter(flights, dep_time < 1200)
```


Complete these lines of code to filter to all major airports in Texas (IAH, DFW, HOU)
```{r filter 2}
flights %>% filter(dest == "IAH", [ADD CODE HERE])
flights %>% filter(dest %in% c("IAH"), [ADD CODE HERE]))
```


## SLICE

Explain what this line of code is doing
```{r slice 1}
flights %>% slice(500:510)
```


Complete these lines of code to view the first 6 rows and the last 12 rows of flight data
```{r slice 2}
flights %>% slice_head([ADD CODE HERE])
flights %>% slice_tail([ADD CODE HERE])
```


## ARRANGE

What is this line of code doing?
```{r arrange 1}
flights %>% arrange(dep_time)
```

How can we alter this code to arrange flights by departure time, largest to smallest?
```{r arrange 2}
flights %>% arrange(desc([ADD CODE HERE]))
```

Write a statement that arranges the flight data by arrival time in descending order, departure time in ascending order, and day in ascending order
```{r arrange3}
flights %>% [ADD CODE HERE]
```


## PRACTICE EXERCISES

Write code to answer each question

1. How many flights departed from JFK on December 3rd?

```{r practice 1.1}
flights %>% filter([ADD CODE HERE])
```

2. How many flights went to ATL in March? What was the shortest flight time?

```{r practice 1.2}
flights %>% [ADD CODE HERE] %>%
            arrange([ADD CODE HERE])
```

3. How many flights are missing arrival delay data?

```{r practice 1.3}
flights %>% [ADD CODE HERE](is.na([ADD CODE HERE]))
```

4. How many flights leaving LGA reported a departure delay over 30 minutes?

```{r practice 1.4}
flights %>% [ADD CODE HERE]
```

5. What was the destination of the flight that had the smallest arrival delay, but traveled the longest distance?

```{r practice 1.5}
flights %>% [ADD CODE HERE]
```


## SELECT

Complete this code to select the origin, destination, and distance columns
```{r select 1}
flights %>% select([ADD CODE HERE])
```

Complete this code to select the columns 13, 6, 2, and 8
```{r select 2}
flights %>% select([ADD CODE HERE]
```

Complete this code to select all columns *except* tail number and time/hour
```{r select 3}
flights %>% select(-[ADD CODE HERE], -[ADD CODE HERE])
```

What is this code doing?
```{r select 4}
flights %>% select(contains("dep"))
```


## MUTATE

What is this code doing?
```{r mutate 1}
flights %>% mutate(speed = distance/air_time*60)
```

Write a line of code that creates a new column called kilometers, which is calculated by multiplying the distance in miles by 1.609
```{r mutate 2}
flights %>% mutate([ADD CODE HERE])
```


## RELOCATE

Complete this line of code to relocate any columns with the string "wind" to appear after the precipitation column
```{r relocate 1}
weather %>% relocate(contains([ADD CODE HERE]), .after = [ADDS CODE HERE])
```

## RENAME

Complete this code to rename the dewp column to dew point
```{r rename 1}
weather %>% rename([ADD CODE HERE] = [ADD CODE HERE])
```


## PRACTICE EXERCISES

1. Make a table that contains data on flight arrivals to SFO from EWR for the 15th day of each month

```{r practice 2.1}
flights %>% filter(origin == [ADD CODE HERE] & dest == [ADD CODE HERE] & [ADD CODE HERE]) %>%
            select(contains([ADD CODE HERE]))
```

2. Make a table of weather observations where the temperature exceeded 90 degrees, but with only origin, date/time, and temp variables

```{r practice 2.2}
weather %>% [ADD CODE HERE] %>%
            select([ADD CODE HERE])
```

3. Make a table of flights that departed early and arrived early, \
  shows only departure delay, arrival delay, and distance with the largest arrival delay first, \
  create a new column called `total_delay` that adds together the departure and arrival delays and     put it in front of the distance column, and \
  rename the departure and arrival delay columns to fully spell out departure/arrival
  
. . . 
  
```{r practice 2.3}
flights %>% filter([ADD CODE HERE]) %>%
            mutate(total_delay = [ADD CODE HERE]) %>%
            select([ADD CODE HERE]) %>%
            rename([ADD CODE HERE]) %>%
            [ADD CODE HERE]
```


## GROUP_BY + SUMMARIZE

What is this code doing?
```{r group summarize 1}
weather %>% group_by(origin, month) %>%
            summarize(mean_temp = mean(temp, na.rm = TRUE))
```

Complete this code to create a table that counts the number of flights from each origin
```{r group summarize 2}
flights %>% group_by([ADD CODE HERE]) %>%
            summarize(count = [ADD CODE HERE])
```


## PRACTICE EXERCISES

1. How many flights departed on the busiest day of the year for each airport?

**[HINT: check out the function `top_n()`]**
```{r practice 3.1}
flights %>% group_by(origin, [ADD CODE HERE]) %>%
            summarize(count = [ADD CODE HERE]) %>%
            group_by([ADD CODE HERE]) %>%
            top_n([ADD CODE HERE])
```

2. What was the coldest day of the year for each airport? How many flights departed each airport on that day?

```{r practice 3.2}
weather %>% group_by([ADD CODE HERE], month, day) %>%
            summarize(min_temp = [ADD CODE HERE]) %>%
            group_by([ADD CODE HERE]) %>%
            filter(min_temp == [ADD CODE HERE])

flights %>% filter(day == [ADD CODE HERE] & month == [ADD CODE HERE]) %>%
            group_by([ADD CODE HERE]) %>%
            summarize(count = [ADD CODE HERE])
```

3. Which destination airport does each NYC airport fly to most often?

```{r practice 3.3}
flights %>% group_by([ADD CODE HERE]) %>%
            summarize([ADD CODE HERE]) %>%
            [ADD CODE HERE]
```

4. What is the total monthly rainfall at each NYC airport in 2023? Which month is the rainiest?

```{r practice 3.4}
[ADD CODE HERE] %>% group_by([ADD CODE HERE]) %>%
                    summarize(total_prcp = [ADD CODE HERE]) %>%
                    pivot_wider(names_from = [ADD CODE HERE], values_from = [ADD CODE HERE])
```

5. What is the longest recorded flight departure delay for each airport?

```{r practice 3.5}
flights %>% [ADD CODE HERE] %>%
            summarize([ADD CODE HERE])
```

6. How many flights departed more than 30 minutes late from each airport? Which carrier was most commonly delayed more than 30 minute?

```{r practice 3.6}
flights %>% [ADD CODE HERE] %>%
            group_by([ADD CODE HERE]) %>%
            summarize([ADD CODE HERE])

flights %>% [ADD CODE HERE] %>%
            group_by([ADD CODE HERE]) %>%
            summarize([ADD CODE HERE]) %>%
            [ADD CODE HERE]
```
