---
title: 'Data Visualization: Group Exercise'
author: "SISMID 2024"
date: ""
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.align = "center")
```

```{r}
library(tidyverse)
data <- read_csv("linelist.csv") |>
  mutate(duration = date_outcome - date_onset)
```


# Part 1: Boxplot

In this lesson we'll be working with the `linelist.csv` dataset. You may have made a boxplot to explore data using base R graphics. That would look like this:
   
```{r boxplot}

boxplot(data$age ~ data$outcome, xlab="Infection Outcome", ylab="Age" ) #instead of assigning x and y values, you can use the syntax y ~ x
```


Remember that each ggplot object has 3 main components:
1. The first argument in the function is the dataframe
2. The second argument takes an aes() function which defines the variables of the dataframe for each dimension of the plot

Let's try running this line of code with just the first 2 components. What do you see?

```{r}
ggplot(data = data, aes(x = duration, y = outcome))
```


3. The third argument tells ggplot what type of graphical geometry we want it to add
Here, we want to make a scatter plot so we'll use the argument geom_point() to visualize the data as points


```{r}
ggplot(data = data, aes(x = duration, y = outcome)) +
  geom_boxplot()
```


Notice that the `+` allows us to add additional functions to the ggplot object.
The layout of the plot suggests a relationship between area and population total.
What if we wanted to reverse this graphical relationship? We could switch the variables supplied to each aesthetic or use `+ coord_flip()`



With additional customization, we can produce a plot with a little more shine.

# Problem 1

Work with your team to create a polished `ggplot2` boxplot.

Use 1-2 of the following graphical arguments in your geom_boxplot function to modify the appearance of the plots. Read the documentation for `geom_boxplot()` to understand how to use them.

* `fill =`
* `color =`
* `alpha =`
* `notch =`
* `outlier.color =`
* `outlier.fill =`
* `outlier.shape =`


Use 2-3 of the following graphical functions to improve your plot (remember to put each on its own line and connect them with a +):


* `xlab()`
* `ylab()`
* `ggtitle()`
* `theme_XXX()` (Try out different themes by placing different themes after the underscore, the full list can be found here: https://ggplot2.tidyverse.org/reference/ggtheme.html)

***Challenge: Modify the theme***

* `theme()` This function lets you set specific components of the theme of a plot. See if you can tweak these to further improve your plot.

```{r Problem 1}

ggplot(data = data, aes(x = duration, y = outcome)) +
  geom_boxplot()
```

 
## Additional Resources:

 * [Cheat Sheet](https://rstudio.github.io/cheatsheets/html/data-visualization.html)
 * [Data Carpentry ggplot2 tutorial](https://datacarpentry.org/R-ecology-lesson/04-visualization-ggplot2.html)
 * A very detailed tutorial is on [ZevRoss](http://zevross.com/blog/2014/08/04/beautiful-plotting-in-r-a-ggplot2-cheatsheet-3/). 
 * A more formal guide is in the [R for Data Science](http://r4ds.had.co.nz/data-visualisation.html) visualization chapter. 

# Part 2: Scatter Plots

We also use scatterplots A LOT to visualize data and communicate our findings. There isn't a "geom_scatter()" function in R. Instead we use `geom_point()`

```{r ggscatter}

data |>
  ggplot(aes(x = ct_blood, y = age))+
  geom_point()
```


## Problem 2: Polished Scatter Plot

1. Create a polished scatterplot of the relationship between age at infection and duration of infection using the function `geom_point()`

2. Add a trendline to the plot using the function `geom_smooth()`.

3. Add x and y labels and a chart title.

***CHALLENGE: Try changing the color of the points to blue and the line to red.***


```{r Problem 2, echo=TRUE}

```

***Challenge: Change the color or shape of the points so that they represent the outcome of infection.***

# Part 3: Bar plots

Unlike boxplots and scatterplots, to create a bar-plot we have to give R summaries of our data (usually means). So we'll combine our dplyr skills with our ggplot skills to make a beautiful barplot!

```{r}
summary_data <- data |>
  group_by(age_cat) |>
  summarize(mean_duration = mean(duration, na.rm = T))

summary_data
```

# Problem 4: Make a bar plot!

use the summary_data to create a boxplot. The geometry for a bar chart of your data values is (somewhat confusingly) `geom_col()`

```{r Problem 4}
summary_data |>

```

***CHALLENGE: Make a bar plot that is grouped by both age category and gender. You'll need to start with a new grouping and summarizing processs***

# Problem 5: Group Challenge

For the final problem you will work to produce a more complicated figure using some of the more advanced/specialized functions of ggplot. Please work in your group to create a polished figure according to the specifications for your group below. You can visualize any relationship in the line-list data, but keep in mind that certain types of plots are better for visualizing some data types and relationships than others.

## Group 1:  Multi-panel Plots

We often want to visualize a relationship between variables in multiple groups side by side. We can facet our data into multiple panels to do this. Create a plot using the linelist dataset that incorporates multiple panels. Use one or more of the following graphical functions:

* `facet_wrap()`
* `facet_grid()`

```{r ggfacet, echo=TRUE, fig.height=10}


```

## Group 2: Histograms and Violin Plots

Histograms are plots of the frequency of a continuous variable in a dataset. They can be useful for visualizing the shape of data distributions.

Violin plots are a stylized version of a boxplot which contain slightly more information about the data distribution. Create a violin plot using the linelist dataset. Use one or more of the following graphical functions:

*`geom_histogram`
*`geom_violin`



```{r gghist, echo=TRUE}



```

## Group 3: Bubble  Plots

Bubble plots

Create a bubble plot using the linelist dataset. Use one or more of the following graphical functions:
* `geom_point()`
* `aes(size = )` (Remember that this function has to go inside another function, see above for examples)
* `scale_size()`


```{r ggbubble, echo=TRUE}



```

## Group 4: Time Series Plots

Time series plots are specialized versions of line/dot plots.

Create a time series plot using the linelist dataset. You may want to use some of your date and tidy data skills to summarize the data by date first. Use one or more of the following graphical functions:

* `geom_line()`


```{r ggtimeseries, echo=TRUE}



```

